\chapter{Problem}
\label{cha:problem}

\sloppy

In the first chapter we presented \textit{template engines} and discussed their theoretical handicaps, in this chapter we will further analyze other handicaps that are presented while using them in a practical setting. This analysis aims to show how fragile the usage of this type of solution can be and the problems that are inherited by using it.

\section{Motivation}
\label{sec:motivation}

Text has evolved with the advance of technology resulting in the creation of \textit{markup languages} \cite{markuplanguages}. Markup languages work by adding annotations to text, the annotations being also known as tags, that allow to add additional information to the text. Each markup language has its own tags and each of those tags add a different meaning to the text encapsulated within them. In order to use markup languages the users can write the text and add all the tags manually, either by fully writing them or by using some kind of text helpers such as text editors with IntelliSense\footnote{\href{https://www.techopedia.com/definition/24580/intellisense}{Intellisense Definition}} which can help diminish the errors caused by manually writing the tags. But even with text helpers the resulting document can violate the restrictions of the respective markup language because the editors don't actually enforce the language rules since there isn't a process similar to a compile process which can either pass of fail. The most that a text editor is highlight the errors to the user.

\noindent
The most common markup language is the \ac{HTML} language, which is heavily used in Web applications. Other uses of the \ac{HTML} language are writing emails, writing reports, etc. The function of \ac{HTML} in Web applications is to define the \textit{user-interface}, which is also known as the \textit{view} of the website. To generate the end view the most common solution is the usage of \textit{template engine}  solutions, which can be considered the \textit{controller} which is responsible for joining the domain data with the views. This approach separates the \textit{view} from the \textit{controller}, which allows both of the different layers of a project to be more independent. But, with the usage of \textit{template engines}, there is another layers of complexity between the \textit{view} and the \textit{controller}, since both of these aspects of the project use different languages. 

\noindent
In the following examples we will present different problems derived from the usage of \textit{template engines} that can be solved by using solutions that have the template incorporated in the programming language. The examples provided in this section use eight different \textit{template engines}: Freemarker\cite{freemarker}, Handlebars\cite{handlebars}, Mustache\cite{mustache}, Pebble\cite{pebble}, Thymeleaf\cite{thymeleaf}, Trimou\cite{trimou}, Velocity\cite{velocity} and Rocker\cite{rocker}. These templates were used to perform benchmarks that will be presented in Chapter \ref{cha:deployment}.

\noindent
We will start by the most basic aspect that we expect from a \ac{XML} document, it should be well formed. Let's start with a very simple example as shown in Listing \ref{lst:wellformedex}.

\lstdefinestyle{problemex}{
  moredelim=**[is][\color{blue}]{@}{@},
  moredelim=**[is][\color{red}]{|}{|},
}

\lstset{language = java}

\bigskip

\begin{minipage}{\linewidth}
\begin{lstlisting}[caption={Badly Formed Document}, label={lst:wellformedex}, style=problemex]
@<html@|>|
	@<body>
		<!-- -->@
	|</body>|
@</html>@
\end{lstlisting}
\end{minipage} 

\noindent
Let's imagine that for some typing mistake the red characters are missing, which means that the opening \texttt{<html>} tag isn't properly closed and that the \texttt{body} element doesn't have a matching closing tag. It would be expected that in the very least \textit{template engine} would issue an error while reading the file at run time. But every one of the \textit{template engines} used with this example haven't issued any kind of error. This is problematic, because the error wasn't caught neither at compile time nor at run time. These kind of errors would only be observable either on a browser or by using any kind of external tool to verify the resulting \ac{HTML} page. By using solutions that incorporate the template within the language this problem doesn't have the possibility of happening since the process of creating tags and properly open and closing them should be performed by the solution and not by the person who is defining the template.

\noindent
The second problem that we are going to point out is the \textit{context objects}. Every \textit{template engine} uses them, since it contains the information that the \textit{template engine} will use to fill out the \textit{placeholders} defined in the textual template file. But what problems do their usage carry? Let's observe the following template in Listing \ref{lst:contextobjs}.

\bigskip

\lstset{language=html, morekeywords={TableElement, String}}

\begin{minipage}{\linewidth}
\begin{lstlisting}[caption={Template with Placeholders}, label={lst:contextobjs}, style=problemex]
<html>
    <body>
        <ul>
        {{#student}}
            <li>
                {{name}}
            </li>
            <li>
                {{number}}
            </li>
        {{/student}}
        </ul>
    </body>
</html>
\end{lstlisting}
\end{minipage} 

\noindent
This template presents a straightforward template, it will receive a \texttt{Student} object that contains a \texttt{name} and \texttt{number} fields. Before going any further let's see how do the template engines pass the information to the template file. Most of the solutions use a simple \texttt{Map<String, Object>}, in this case a correct \textit{context object} should look like the \texttt{Map} object created in Listing \ref{lst:contextobj}.

\bigskip

\lstset{language=java, morekeywords={Map, String, Object, HashMap, put, Student}}

\begin{minipage}{\linewidth}
\begin{lstlisting}[caption={Correct Context Object}, label={lst:contextobj}]
Map<String, Object> context = new HashMap<>();
context.put("student", new Student("Luis", 39378));
\end{lstlisting}
\end{minipage} 

\noindent
The previous example makes sense, there is one object in the \textit{context object} with the \texttt{student} key with an instance of a \texttt{Student} object, which contains a \texttt{name} and \texttt{number} fields, which corresponds with the usage performed in the previously defined template in Listing \ref{lst:contextobjs}. But what if instead of that context object we used another such as Listing \ref{lst:contextobj1} or Listing \ref{lst:contextobj2}?

\bigskip

\lstset{language=java, morekeywords={Map, String, Object, HashMap, put, Student}}

\begin{minipage}{\linewidth}
\begin{lstlisting}[caption={Context Object With Errors - 1}, label={lst:contextobj1}]
Map<String, Object> context = new HashMap<>();
context.put("teacher", new Student("Luis", 39378));
\end{lstlisting}
\end{minipage} 

\bigskip

\lstset{language=java, morekeywords={Map, String, Object, HashMap, put, Teacher}}

\begin{minipage}{\linewidth}
\begin{lstlisting}[caption={Context Object With Errors - 2}, label={lst:contextobj2}]
Map<String, Object> context = new HashMap<>();
context.put("student", new Teacher("MEIC", "ADDETC"));
\end{lstlisting}
\end{minipage} 

\noindent
Both of those \textit{context objects} have problems. The first one, Listing \ref{lst:contextobj1}, has a wrong key, \texttt{teacher}, when the template is expecting an object with the \texttt{student} key. The second one has the right key but has a different type, which doesn't contain the fields that are expected. So, with these information in mind how will the eight template engines react when receiving these two wrongly defined \textit{context objects}? The Rocker template engine is the one which deals with it the best since it defines in the template the type that will receive and since his template file is converted in a Java class at compile time its usages are all safe regarding \textit{context object} since the Java compiler validates if the object received as \textit{context object} matches the expected type. The remaining seven more classical solutions are more fragile, none of them issue any compile time warning. As for run time errors, only Freemarker issues an exception with a similar example to Listing \ref{lst:contextobj1} and in the second case, Listing \ref{lst:contextobj2}, only Freemarker and Thymeleaf throw an exception. The remaining solutions ignore the fact that something that is expected isn't there and delay the error finding process until the generated file is manually validated. By using templates that are defined within the language the \textit{context objects} can be typified and with the help of the Java compiler all of these errors can be found at compile time.

\noindent
While the previous examples approach most of the practical uses of \textit{template engines}, there are other quality of life aspects that derive from the usage of templates defined within the programming language. The main improvement is that the syntax is always the same, there is no need for additional syntaxes. For example, even for the simplest templates we have to atleast deal with three syntaxes. In the following example we will use the Pebble \textit{template engine}, one who requires less verbose. In this example we define a template to write an \ac{HTML} document that presents the \texttt{name} of all the \texttt{Student} objects present in a \texttt{Collection} of \texttt{Student} as shown in Listing \ref{lst:listofnamestemplate}.

\bigskip

\lstset{language=html, morekeywords={TableElement, String}}

\begin{minipage}{\linewidth}
\begin{lstlisting}[caption={List of Student Names Template - Pebble}, label={lst:listofnamestemplate}]
<html>
	<body>
	    <ul>
			{% for student in students %}
		  	<li>{{student.name}}</li>
			{% endfor %}
		</ul>  
  	</body>
</html>
\end{lstlisting}
\end{minipage} 

\noindent
In this template alone we need to use two distinct syntaxes, the \ac{HTML} language and the Pebble syntax to indicate that the template will receive a \texttt{Collection} that should be iterated  and create \texttt{li} tags containing the \texttt{name} field of the received type. Apart from the template definition we also need the Java code to generate the complete document, as shown in Listing \ref{lst:listofnamesjava}.

\bigskip

\lstset{language=java, morekeywords={PebbleEngine, Builder, build, getTemplate, StringWriter, Map, String, Object, HashMap, getListOfStudents, evaluate, toString, put}}

\begin{minipage}{\linewidth}
\begin{lstlisting}[caption={List of Student Names Java - Pebble}, label={lst:listofnamesjava}]
PebbleEngine engine = new PebbleEngine.Builder().build();
template = engine.getTemplate("templateName.html");
StringWriter writer = new StringWriter();

Map<String, Object> context = new HashMap<>();
context.put("students", getListOfStudents());

template.evaluate(writer, context);

String document = writer.toString();
\end{lstlisting}
\end{minipage} 

\noindent
Even though the template in Listing \ref{lst:listofnamestemplate} is simple the usage of multiple syntaxes introduces more complexity to the problem. If we escalate the complexity of the template and the number of different types used in the context object mistakes are bound to happen, which would be fine if the template engines gave any kind of feedback on errors, but we already shown that most errors aren't reported. Let's take a peek of how this same template would be presented in the latest HtmlFlow version, shown in Listing \ref{lst:htmlflowlistofnames}.

\bigskip

\lstset{language=java, morekeywords={String, view, DynamicHtml, CurrentClass, render, getStudentList, Iterable, Student, html, body, ul, dynamic, forEach, text, li, getName}}

\begin{minipage}{\linewidth}
\begin{lstlisting}[caption={List of Student Names - HtmlFlow/xmlet}, label={lst:htmlflowlistofnames}, literate={º}{\textdegree}1]
String document = DynamicHtml.view(CurrentClass::studentListTemplate)
                             .render(getStudentsList());
    
static void studentListTemplate(DynamicHtml<Iterable<Student>> view,                                
                                Iterable<Student> students){
    view.html()
            .body()
                .ul()
                    .dynamic(ul -> 
                      students.forEach(student ->                   
                        ul.li().text(student.getName()).º()))
                .º()
            .º()
        .º();
}
\end{lstlisting}
\end{minipage} 

\noindent
With this solution we have a very compact template definition, where the context object, i.e. the \texttt{Iterable<Student> students}, is validated by the Java compiler in compile time which guarantees that any document generated by this solution will be valid since the program wouldn't compile otherwise. This solution internally guarantees that the \ac{HTML} tags are created properly, having matching opening and ending tags, meaning that every document generated by this solution will be well formed regardless of the defined template. 

\noindent
Another quality of life improvement that we obtain by using the templates defined within the language is navigability. One aspect that is very common in template engines solutions is to define \textit{partial views} that can be reused in different views, but with regular text editors sometimes is hard to navigate back and forth between partial/regular templates. By using templates within the language we are able to quickly move between templates, since the template is either a method or a field and most \ac{IDE}s allow to quickly access both of them.

\section{Problem Statement}
\label{sec:problemstatement}

The problem that's being presented revolves around the handicaps of template engines, the lack of compilation of the language used within the template, the performance overhead that it introduces and the issues that it was when the complexity increases, as it was presented in the previous Section \ref{subsec:templateengineshandicaps}. To tackle those handicaps we suggested the automated generation of a strongly typed fluent interface. In order to show how that fluent interface will effectively work we will now present a small example which consists on the \texttt{html} element, Listing \ref{lst:codegenerationexample}, described in \ac{XSD} of the \ac{HTML}5 language definition. The presented example is simplified for explanation purposes.

\bigskip

\lstset{
	language=XML,
	morekeywords={xs:element, name, xs:complexType, xs:choice, ref, xs:attributeGroup, xs:attribute, type}
}

\begin{minipage}{\linewidth}
\begin{lstlisting}[caption={Code Generation XSD Example},captionpos=b, ,label={lst:codegenerationexample}]
<xs:attributeGroup name="commonAttributeGroup">
    <xs:attribute name="someAttribute" type="xs:string">
</xs:attributeGroup>

<xs:element name="html">
    <xs:complexType>
        <xs:choice>
            <xs:element ref="body"/>
            <xs:element ref="head"/>
        </xs:choice>
        <xs:attributeGroup ref="commonAttributeGroup" />
        <xs:attribute name="manifest" type="xs:string" />
    </xs:complexType>
</xs:element>
\end{lstlisting}
\end{minipage}

\noindent
With this example there is a multitude of classes that need to be created, apart from the always present supporting infrastructure that will be presented in Section \ref{sec:supportinginfrastructure}. 

\begin{itemize}
	\item Html Element - A class that represents the \texttt{Html} element (Listing \ref{lst:htmlclass}), deriving from \texttt{AbstractElement}.
	\item Body and Head Methods - Both methods present in the \texttt{Html} class (Listing \ref{lst:htmlclass}) that add \texttt{Body} (Listing \ref{lst:bodyclass}) and \texttt{Head} (Listing \ref{lst:headclass}) instances to \texttt{Html} children list.
	\item Manifest Method - A method present in \texttt{Html} class (Listing \ref{lst:htmlclass}) that adds an instance of the \texttt{Manifest} attribute (Listing \ref{lst:manifestattributeclass}) to the \texttt{Html} attribute list.
\end{itemize}

\lstset{language=Java, morekeywords={AbstractElement, CommonAttributeGroup, Body, addChild, Head, String, AttrManifest, addAttr}}

\begin{minipage}{\linewidth}
\begin{lstlisting}[caption={Html Element Class},captionpos=b,label={lst:htmlclass}]
class Html extends AbstractElement implements CommonAttributeGroup {
    public Html() { }
    
    public void accept(Visitor visitor){
		visitor.visit(this);    
    }
    
    public Html attrManifest(String attrManifest) {
        return this.addAttr(new AttrManifest(attrManifest));
    }
    
    public Body body() { return this.addChild(new Body()); }
        
    public Head head() { return this.addChild(new Head()); }
}
\end{lstlisting}
\end{minipage}

\begin{itemize}
	\item Body and Head classes - Classes for both \texttt{Body} (Listing \ref{lst:bodyclass}) and \texttt{Head} (Listing \ref{lst:headclass}) elements, similar to the generated \texttt{Html} class (Listing \ref{lst:htmlclass}). The class contents will be dependent on the contents present in the concrete \texttt{xsd:element} nodes.
\end{itemize}

\bigskip

\begin{minipage}{\linewidth}
\begin{lstlisting}[caption={Body Element Class},captionpos=b,label={lst:bodyclass}]
public class Body extends AbstractElement {
    //Similar to Html, based on the contents of the respective
    //xsd:element node.
}
\end{lstlisting}
\end{minipage}

\bigskip

\begin{minipage}{\linewidth}
\begin{lstlisting}[caption={Head Element Class},captionpos=b,label={lst:headclass}]
public class Head extends AbstractElement {
    //Similar to Html, based on the contents of the respective 
    //xsd:element node.
}
\end{lstlisting}
\end{minipage}

\begin{itemize}
	\item Manifest Attribute - A class that represents the \texttt{Manifest} attribute  (Listing \ref{lst:manifestattributeclass}), deriving from \texttt{BaseAttribute}.
\end{itemize}

\bigskip

\lstset{language=Java, morekeywords={String, BaseAttribute}}

\begin{minipage}{\linewidth}
\begin{lstlisting}[caption={Manifest Attribute Class},captionpos=b,label={lst:manifestattributeclass}]
public class AttrManifestString extends BaseAttribute<String> {
    public AttrManifestString(String attrValue) {
        super(attrValue);
    }
}
\end{lstlisting}
\end{minipage}

\begin{itemize}
	\item CommonAttributeGroup Interface - An interface with default methods that add the group attributes to the concrete element (Listing \ref{lst:commonattributegroup}).
\end{itemize}

\bigskip

\lstset{language=Java, morekeywords={addAttr, Html, Element, String, SomeAttribute}}

\begin{minipage}{\linewidth}
\begin{lstlisting}[caption={CommonAttributeGroup Interface},captionpos=b,label={lst:commonattributegroup}]
public interface CommonAttributeGroup extends Element {
    default Html attrSomeAttribute(String attributeValue) {
        this.addAttr(new SomeAttribute(attributeValue));
        return this;
    }
}
\end{lstlisting}
\end{minipage}

\noindent
By analyzing this little example we can observe how the \texttt{xmlet} solution implements one of its most important features that was lacking in the \textit{template engine} solutions, the user is only allowed to generate a tree of elements that follows the rules specified in the \ac{XSD} file of the given language, e.g. the user can only add \texttt{Head} and \texttt{Body} elements as children to the \texttt{Html} element and the same goes for attributes as well, the user can only add a \texttt{Manifest} or \texttt{SomeAttribute} objects as attribute. This solution effectively uses the Java compiler to enforce the specific language restrictions, most of them at compile time. The other handicaps are also solved, the template can now be defined within the Java language eradicating the necessity of textual files that still need to be loaded into memory and resolved by the \textit{template engine}. The complexity and flexibility issues are also tackled by moving all the parts of the problem to the Java language, it removes the necessity of additional syntax and now the Java syntax can be used to create the templates.

\newpage

\section{Approach}
\label{sec:approach}

The approach to achieve a solution was to divide the problem into three distinct aspects, as previously stated in Section  \ref{sec:thesisstatement}. 

\noindent
The XsdParser project will be an utility project which is needed in order to parse all the external \ac{DSL} rules present in the \ac{XSD} document into structured Java classes. 

\noindent
The XsdAsm is the most important aspect of the \texttt{xmlet} solution, since it is the aspect which will deal with the generation of all the \textit{bytecodes} that make up the classes of the Java \textit{fluent interface}. This project should translate as many rules of the parsed language definition, its \ac{XSD} file, into the Java language in order to make the resulting \textit{fluent interface} as much similar as possible to the language definition.

\noindent
The HtmlApi will be a representation of client aspect of \texttt{xmlet} solution. It is a concrete client of the XsdAsm project, it will use the \ac{HTML}5 language definition file in order to request of XsdAsm a strongly typed \textit{fluent interface}, named HtmlApi. This use case is meant to be used by the HtmlFlow library which will use HtmlApi to manipulate the \ac{HTML} language to write well formed documents.