\chapter{Deployment}
\label{cha:deployment}

\section{Github Organization} % (fold)
\label{sec:github}

This project and all its components belong to a github organization called xmlet
%TODO url ou footnote para o link do repositório.
. The aim of that organization is to contain all the related projects to this dissertation. All the generated APIs are also created as if they belong to this organization. 

\section{Maven} % (fold)
\label{sec:maven}

In order to manage the developed projects a tool for project organization and deployment was used, named Maven. Maven has the goal of organizing a project in many different ways, such as creating a standard of project building, managing project dependencies. Maven was also used to generate documentation and deploying the projects to a central code repository, Maven Central Repository
%TODO footnote com url 
.

\section{Sonarcloud} % (fold)
\label{sec:sonarcloud}

Code quality and its various implications such as security, low performance and bugs should always be an important issue to a programmer. With that in mind all the projects present in this dissertation were evaluated in various metrics and the results made public for consultation. This way, either future users of those projects or just for the developers, this metrics can be used to serve as another way of validating the quality of the produced code. The tool to perform this evaluation was Sonarcloud
%TODO footnote para a dashboard da página da organização xmlet.
, which provides free of charge evaluations and stores the results which are available for everyone.

\section{Testing metrics} % (fold)
\label{sec:testingmetrics}

Perform efficiency tests comparing the HtmlApi, j2html and Apache Velocity. The test should be based on a html page with multiple elements and attributes, probably the test should be performed with different number of html elements, like 10, 50, 100, 1000.